---
title: "run_analysis"
author: "Stephanie R. Beck Roth"
date: "July 21, 2015"
output: html_document
---
### This program is written to fulfill the requirements of
### the Getting and Cleaning Data Coursera Course. Smartphone
### Data was extracted from the following websites belonging
### to UCI Machine Learning Repository.
 "http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones#"
### Program written by Stephanie R. Beck Roth
### Due 7-26-2015

# There are 2 parts to this CodeBook.

## 1: The first is the code to produce the tidytable.txt file and the second is a description of the variables inclcuded in the text file.

### The following is the R code to Load and Produce the alldatatrim master tidy dataset as required from objective 4.
```{r, echo=TRUE}
library(doBy)
library(dplyr)
library(tidyr)
library(gtools)
#Get Data if not already downloaded to save time.  :)

if (!file.exists("./UCI\ HAR\ Dataset/test")){
    fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
    download.file(fileUrl,destfile = "./SmartPhone.zip", method="curl")
    mycsv <- unzip("./SmartPhone.zip", exdir = ".")
}

# The first 5 text files are descriptors of the data.
# Files 5-13 contain Inertial Signals tests data.
# Files 14-16 contain subject, X and Y tests data, respectively.
# Files 17-25 contain the Inertial Signals training data.
# Files 26-28 contain subject, X and Y training data, respectively.
# Only including the following files for analysis per discussion forum post by course TA, 
# David Hood at https://class.coursera.org/getdata-030/forum/thread?thread_id=37

# Read file containing activity labels for dytest and dytrain activity codes
aln <- read.delim("./UCI\ HAR\ Dataset/activity_labels.txt", sep=" ",
                  col.names=c("code","activitylabel"),header=FALSE)

dytest <- read.table("./UCI\ HAR\ Dataset/test/y_test.txt", 
                     col.names = "code", header=FALSE)
dytrain <- read.table("./UCI\ HAR\ Dataset/train/y_train.txt", 
                      col.names = "code", header=FALSE)

# Read file containing column headers for dxtest and dxtrain
coln <- read.delim("./UCI\ HAR\ Dataset/features.txt", sep=" ",header=FALSE)
cnames <- as.character(coln$V2)

# Read file containing participant identififiers for dstest 
# Note: There are only 30 participants and each participant is either in the
# training dataset or the test dataset.
dstest <- read.delim("./UCI\ HAR\ Dataset/test/subject_test.txt",header=FALSE)
part_id_test <- as.factor(dstest$V1)

# Read test Content Files and wrap the data frames as data tables and setting variables to those in features.txt
dxtest <- tbl_df(read.table("./UCI\ HAR\ Dataset/test/X_test.txt", col.names=cnames))
# Create testdata which will have the following new variables stored in the beginning of the table dxtest
# dxtest$set <- "test"
# dxtest$part_id <- part_id_test
# dxtest$code <- dytest$code
testdata <- cbind(part_id=part_id_test,set="test",code=dytest$code,dxtest)

# Read file containing participant identififiers for dstrain 
dstrain <- read.delim("./UCI\ HAR\ Dataset/train/subject_train.txt",header=FALSE)
part_id_train <- as.factor(dstrain$V1)

# Read train Content Files and wrap the data frames as data tables and setting variables to those in features.txt
dxtrain <- tbl_df(read.table("./UCI\ HAR\ Dataset/train/X_train.txt", col.names=cnames))

# Create traindata which will have the following new variables stored in the beginning of the table dxtrain
# dxtrain$set <- "train"
# dxtrain$part_id <- part_id_train
# dxtrain$code <- dytrain$code
traindata <- cbind(part_id=part_id_train,set="train",code=dytrain$code,dxtrain)

# Create Data set with both test and train data
# Using rbind since the variable names are identical
alldata <- rbind(testdata,traindata)
# Introducde the activitylabel variable which exlains the information in the variable code.
alldata <- merge(alldata, aln, by="code")
# Reorder the table so that the activity label variable is near the beginning in the dataset
alldata <- alldata[, c(1,2,3,565,4:564)]
# Deleting the code variable to make it tidy since the activitylabel variable is now in the dataset
# to convey the same information.
alldata <- select(alldata, -code)
# Select only part_id, set, activitylabel (columns 1-3)
# and those variables containing mean or std in their names.
cols <- 1:3
colsm <- grep("mean", colnames(alldata))
cols <- append(cols,colsm)
colstd <- grep("std", colnames(alldata))
cols <- sort(append(cols,colstd))
alldatatrim <- alldata[,cols]
```

### The following R code creates the second independent tidy data set containing the average of each variable for each activity and each subject.

```{r, echo=TRUE}
# Creating the tidy dataset
tidytable <-summaryBy(. ~set+part_id+activitylabel,data=alldatatrim,FUN=mean)
```
### The following R code writes the second independent tidy data set containing the average of each variable for each activity and each subject.

```{r, echo=TRUE}
# Writing the tidy dataset
write.table(tidytable, file="./tidytable.txt", row.name=FALSE)
```

# 2:  This begins the second portion of the CodeBook containing the variable descriptions.

### The following variables were created and extracted from the data contained in the zip file:
 https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip
### Further documentation on the collection of the original variables can be found within the README.txt of the data which upon extraction will reside within a directory labeled "UCI HAR Dataset".  This original README.txt file provides the necessary details regarding the specifics of the data collection methodologies from which the following variables were created and should be referenced prior to interpretation of the following information for a more thorough and detailed understanding.

### This dataset contains summary measurements on the 30 participants who were recruited into the two distinct groups, training and test, to perform the original analysis.  Thus, every  participant is a member of only one dataset. 

####     The variable set contains a 2 level factor containing test and train indicate data belonging  to the test dataset and the training dataset, respective.

####     "part_id" contains a 30 level factor comprised of integers in the 1-30 range. This variable indicates the original participant identification number. 
 
####     "activitylabel" contains one of the 6 activity labels: "LAYING", "SITTING", "STANDING", "WALKING", "WALKING_UPSTAIRS", "WALKING_DOWNSTAIRS".

####     The remaining 79 variables contain the mean of their respective measurements from the original datasets calculated separately by the participant and activity labels.  These summaries exist only for the original measurements containing mean and standard deviation. 
#### Note:
####     The run_analysis.R code also indicates the means were performed by the set variable but as no participant is a member of both the training and test datasets, it merely provides a more thorough grouping summary and has no bearing on the calculations.) 

####     Per the original documentation found in the README.txt. "The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

####    The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain. See 'features_info.txt' for more details. " 

####    Credit:(Jorge L. Reyes-Ortiz, Davide Anguita, Alessandro Ghio, Luca Oneto; Smartlab - Non Linear Complex Systems Laboratory)

####     The 79 variables following set, part_id and activitylabel contain the same name as the original data variables where each -, ( and ) are replaced by a . from the original variable and conclude with ".mean" to indicate the mean of the original variable was taken and will have the same units as their original measurements described above.

```{r, echo=FALSE}
names(tidytable)
```

```{r, echo=FALSE}
# Copy the .md file to 
b <- file.copy(from="run_analysis.md", to="CodeBook.md")
```


